---
title: "GLMM rCCR vs. tempertaure"
author: "Laura Pettrich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set up Packages

```{r}
# Clean environment
rm(list = ls())

# Load libraries
library(tidyverse)
library(lme4)    # For mixed-effects modeling
library(lmerTest) # For p-values and significance testing
library(raster)
library(scales)
library(reshape2)
library(broom.mixed)
library(glmmTMB)
library(DHARMa)
library(mgcv)

# Set directory
setwd("~/sciebo/RecombinationLandscape_CRIP/01_Revison/NewResults/")
getwd()
```

Read in relevant data

```{r}
############
# 1. MSMC2 #
############
# B) CROSS-COALESCENCE RATE
#---------------------------
# Read in data
# Combined Cross-Coalescence
mfmg <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MF1-MG5_msmc2.final.txt", header = TRUE)
mfnmf <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MF1-NMF4_msmc2.final.txt", header = TRUE)
mfsi <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MF1-SI4_msmc2.final.txt", header = TRUE)
mfss <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MF1-SS4_msmc2.final.txt", header = TRUE)
mgnmf <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MG2-NMF4_msmc2.final.txt", header = TRUE)
mgsi <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MG2-SI4_msmc2.final.txt", header = TRUE)
mgss <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_MG2-SS4_msmc2.final.txt", header = TRUE)
nmfsi <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_NMF1-SI4_msmc2.final.txt", header = TRUE)
nmfss <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_NMF1-SS4_msmc2.final.txt", header = TRUE)
siss <- read.table("~/sciebo/RecombinationLandscape_CRIP/01_Revison/Sorted_final_files/Final_Supplement/Zenodo/MSMC2/combined_SI1-SS4_msmc2.final.txt", header = TRUE)


# determine mutation rate and generation time
mu <- 4.27*10^-9    #from Waldvogel & Pfenninger 2021; mutation rate
gen_mf <- 1/9.07    # generation time = time/generations
gen_mg <-  1/7.85 
gen_nmf <- 1/7.7
gen_si <- 1/10.57
gen_ss <- 1/14.86

# to get relative gene flow, you can compute the relative cross-coalescence rate: 2 * lambda01 / (lambda00 + lambda11)

str(mfmg)
mfmg <- mfmg %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mfnmf <- mfnmf %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mfsi <- mfsi %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mfss <- mfss %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mgnmf <- mgnmf %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mgsi <- mgsi %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

mgss <- mgss %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

nmfsi <- nmfsi %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

nmfss <- nmfss %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))

siss <- siss %>% 
  mutate(rel.cc = (2*lambda_01)/(lambda_00+lambda_11))



# add column with pop.size, generations and years
mfmg <- mfmg %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),    # generations ago
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mf+gen_mg)/2)))   # multiple by generation time to get years

mfnmf <- mfnmf %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),    
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mf+gen_nmf)/2)))   

mfsi <- mfsi %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),    
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mf+gen_si)/2)))   

mfss <- mfss %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),    
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mf+gen_ss)/2)))   

mgnmf <- mgnmf %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mg+gen_nmf)/2)))

mgsi <- mgsi %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mg+gen_si)/2)))

mgss <- mgss %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_mg+gen_ss)/2)))

nmfsi <- nmfsi %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_nmf+gen_si)/2)))

nmfss <- nmfss %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_nmf+gen_ss)/2)))

siss <- siss %>% 
  mutate(generations = (((left_time_boundary+right_time_boundary)/2)/mu),
         years = (((left_time_boundary+right_time_boundary)/2)/mu*((gen_si+gen_ss)/2)))



# remove outliers in data
mfmg <- mfmg[6:31,]
mfnmf <- mfnmf[6:31,]
mfsi <- mfsi[6:31,]
mfss <- mfss[6:31,]
mgnmf <- mgnmf[6:31,]
mgsi <- mgsi[6:31,]
mgss <- mgss[6:31,]
nmfsi <- nmfsi[6:31,]
nmfss <- nmfss[6:31,]
siss <- siss[6:31,]

# merge data frames
mdata <- rbind(mfmg,mfnmf,mfsi,mfss,mgnmf,mgsi,mgss,nmfsi,nmfss,siss)

# add column with population names
mdata <- data.frame(mdata,Population = rep(c("MF-MG","MF-NMF","MF-SI","MF-SS",
                                             "MG-NMF","MG-SI","MG-SS",
                                             "NMF-SI","NMF-SS","SI-SS"),
                                           times=c(nrow(mfmg),nrow(mfnmf),nrow(mfsi),nrow(mfss),nrow(mgnmf),nrow(mgsi),nrow(mgss),
                                                   nrow(nmfsi),nrow(nmfss),nrow(siss))))


head(mdata)

# Check data 
ggplot(mdata, aes(x = years, y = rel.cc, colour = Population)) + 
  geom_step(size=0.75)  +
  scale_x_continuous(limits=c(10^2,10^6), trans="log10") +
  scale_y_continuous(limits = c(0,1.2), breaks = c(0, 0.5, 1)) +
  labs(x = "Years ago", y = "Relative \ncross-coalescence \nrate") + 
  guides(color = guide_legend(nrow = 2, byrow = FALSE, title= "Population\npair"))

#-------------------------------------------------------------------------------------------------------------
###########################
# 2. CHELSA CLIMATE MODEL #
###########################
# It's better to use annual mean temperature https://chelsa-climate.org/chelsa-trace21k/
# bio1 = Annual Mean Temperature [°C]
# kBP 1-22
# Read in tif 
setwd("~/Documents/03_Master/Thesis/ClimateData/CHELSA/bio1") #linux computer

rlist=list.files(getwd(), pattern=".tif", full.names=FALSE)
for(i in rlist) { assign(unlist(strsplit(i, "[.]"))[1], raster(i)) } 

kBP1 <- raster("CHELSA_TraCE21k_bio01_10_V1.0.tif")
kBP2 <- raster("CHELSA_TraCE21k_bio01_0_V1.0.tif")
kBP3 <- raster("CHELSA_TraCE21k_bio01_-10_V1.0.tif")
kBP4 <- raster("CHELSA_TraCE21k_bio01_-20_V1.0.tif")
kBP5 <- raster("CHELSA_TraCE21k_bio01_-30_V1.0.tif")
kBP6 <- raster("CHELSA_TraCE21k_bio01_-40_V1.0.tif")
kBP7 <- raster("CHELSA_TraCE21k_bio01_-50_V1.0.tif")
kBP8 <- raster("CHELSA_TraCE21k_bio01_-60_V1.0.tif")
kBP9 <- raster("CHELSA_TraCE21k_bio01_-70_V1.0.tif")
kBP10 <- raster("CHELSA_TraCE21k_bio01_-80_V1.0.tif")
kBP11 <- raster("CHELSA_TraCE21k_bio01_-90_V1.0.tif")
kBP12 <- raster("CHELSA_TraCE21k_bio01_-100_V1.0.tif")
kBP13 <- raster("CHELSA_TraCE21k_bio01_-110_V1.0.tif")
kBP14 <- raster("CHELSA_TraCE21k_bio01_-120_V1.0.tif")
kBP15 <- raster("CHELSA_TraCE21k_bio01_-130_V1.0.tif")
kBP16 <- raster("CHELSA_TraCE21k_bio01_-140_V1.0.tif")
kBP17 <- raster("CHELSA_TraCE21k_bio01_-150_V1.0.tif")
kBP18 <- raster("CHELSA_TraCE21k_bio01_-160_V1.0.tif")
kBP19 <- raster("CHELSA_TraCE21k_bio01_-170_V1.0.tif")
kBP20 <- raster("CHELSA_TraCE21k_bio01_-180_V1.0.tif")
kBP21 <- raster("CHELSA_TraCE21k_bio01_-190_V1.0.tif")
kBP22 <- raster("CHELSA_TraCE21k_bio01_-200_V1.0.tif")

# Create a data.frame with sample site coordinates
site <- c("MG","NMF", "MF", "SI", "SS")
lon <- c(9.0819270 , 6.2156670, 4.8865000 , 8.3473320, -4.5267980)
lat <- c(50.1680610, 49.1765430, 45.8616760, 45.4036180, 37.399080)
samples <- data.frame(site, lon, lat, row.names="site")
samples

# Extract data from tif for your sites
ID <- seq(-200, 10, by=10)
d <- c(paste("CHELSA_TraCE21k_bio01_", ID, "_V1", sep = ""))


temp.data <- samples 
temp.data$kBP1 <- raster::extract(kBP1, samples) # 1kBP
temp.data$kBP2 <- raster::extract(kBP2, samples)  # 2kBP
temp.data$kBP3 <- raster::extract(kBP3, samples) # 3kBP
temp.data$kBP4 <- raster::extract(kBP4, samples) # 4kBP
temp.data$kBP5 <- raster::extract(kBP5, samples) # 5kBP
temp.data$kBP6 <- raster::extract(kBP6, samples) # 6kBP
temp.data$kBP7 <- raster::extract(kBP7, samples) # 7kBP
temp.data$kBP8 <- raster::extract(kBP8, samples) # 8kBP
temp.data$kBP9 <- raster::extract(kBP9, samples) # 9kBP
temp.data$kBP10 <- raster::extract(kBP10, samples) # 10kBP
temp.data$kBP11 <- raster::extract(kBP11, samples) # 11kBP
temp.data$kBP12 <- raster::extract(kBP12, samples) # 12kBP
temp.data$kBP13 <- raster::extract(kBP13, samples) # 13kBP
temp.data$kBP14 <- raster::extract(kBP14, samples) # 14kBP
temp.data$kBP15 <- raster::extract(kBP15, samples) # 15kBP
temp.data$kBP16 <- raster::extract(kBP16, samples) # 16kBP
temp.data$kBP17 <- raster::extract(kBP17, samples) # 17kBP
temp.data$kBP18 <- raster::extract(kBP18, samples) # 18kBP
temp.data$kBP19 <- raster::extract(kBP19, samples) # 19kBP
temp.data$kBP20 <- raster::extract(kBP20, samples) # 20kBP
temp.data$kBP21 <- raster::extract(kBP21, samples) # 21kBP
temp.data$kBP22 <- raster::extract(kBP22, samples) # 22kBP

#fix("temp.data")

# new data fram with only temperature
temp <- as.data.frame(t(temp.data[,-c(1:2)]))
temp$kBP <- c(1:22)


d <- melt(temp, id.vars="kBP")

d$variable <- factor(d$variable ,                 # Relevel group factor
                     levels = c("MG", "NMF", "MF", "SI", "SS"))

d$kBPtrans <- d$kBP*1000 # Transform kBP to BP

str(d)
temperature_data <- d %>%
  rename(
    Population = variable,
    temperature = value,
    time = kBPtrans 
  )


head(temperature_data)
```

Interpolate missing data

```{r}
########################################
# 3. INTERPOLATE rCCR                  #
########################################
# time point for temperature and rCCR  are different
# I need to interpolate the more frequent  values (rCCR ) to fit to the less frequent (temperature)
full_time_sequence <- data.frame(
  time = seq(1000, 22000, by = 1000)  # 1000-year intervals
)

# Rename mean years to time
head(mdata)
colnames(mdata)[colnames(mdata) == "years"] <- "time"



library(zoo)

# Perform interpolation within each location
interpolated_data <- mdata %>%
  group_by(Population) %>%
  complete(time = seq(1000, 22000, by = 1000)) %>%  # Ensure complete time sequence per location
  arrange(Population, time) %>%                      # Arrange by location and time
  mutate(
    interpolated = is.na(rel.cc), # Flag the interpolated values (NA before interpolation)
    rel.cc = zoo::na.approx(rel.cc, na.rm = FALSE)) %>%  # Interpolate rel.cc
  ungroup()

# View the interpolated data
head(interpolated_data)

interpolated_data <- interpolated_data %>%
  dplyr::select(
    Population, time, time_index, rel.cc, interpolated
  )

# Inspect data
scientific_10 <- function(x) {   parse(text=gsub("e\\+*", " %*% 10^", 
                                                 scales::scientific_format()(x))) }

# original data
ggplot(mdata, aes(x = time, y = rel.cc, colour = Population)) + 
  geom_step(size=0.75)  +
  scale_x_continuous(limits=c(10^2,10^6), trans="log10", labels = scientific_10) +
  scale_y_continuous(labels = comma) 

# interpolated data
ggplot(interpolated_data, aes(x = time, y = rel.cc, color = Population)) +
  geom_step(size=0.75) +
  geom_point(data = mdata, aes(x = time, y = rel.cc), size = 2) +
  geom_point(data = interpolated_data %>% filter(interpolated == TRUE), 
             aes(x = time, y = rel.cc), color = "red", size = 2) +
  #xlim(0, 22000) +
  labs(title = "Interpolated rCCR Over Time by Location",
       x = "Time (years ago)",
       y = "Ne") +
  scale_x_continuous(limits=c(10^2,10^6), trans="log10", labels = scientific_10) +
  scale_y_continuous(labels = comma) 
```

Merge data frames

```{r}
########################################
# 3. MERGE DATAFRAMES                  #
########################################
# Should have columns
#  time for the time points 
#  rel.cc
#  temperature for temperature value
#  location for population location

# structure of dataframe needs to change, so that populations can be distinguished
interpolated_data <- interpolated_data %>%
  mutate(PopulationPair = Population)  # Create a new PopulationPair column

interpolated_data <- interpolated_data %>%  
  separate(Population, into = c("Population1", "Population2"), sep = "-") %>%  # Split Population into two populations
  pivot_longer(cols = c(Population1, Population2), names_to = "PopTag", values_to = "Population") 


interpolated_data <- interpolated_data %>%
  dplyr::select(PopulationPair, time, time_index, rel.cc, interpolated, Population)  # Keep PopulationPair and relevant columns

# View the resulting dataframe
print(interpolated_data)


# join with temperature
data <- left_join(interpolated_data, temperature_data, by = c("Population", "time"))

head(data)

# Select 1000 to 22000 years
data <- data %>%
  filter(time >= 1000) %>%
  filter(time <= 22000) %>%
  mutate(
    time = (time) # *(-1)) - mean(time*(-1))) / sd(time*(-1))  # Center and scale, years go into past so they should be negative
  )
  #drop_na(temperature)


head(data)
setwd("~/sciebo/RecombinationLandscape_CRIP/01_Revison/NewResults/")
write.table(data, file = "rCCR_interpolated_temperature_combined.csv", sep = ",")

# every population pair has now two temperature values per time point
```

Generalized Linear Mixed Model (GLMM)

```{r}
###########################################
# 4. Generalized Linear Mixed Model (GLMM)#
###########################################
#----------------

# Fit different GLMMs with appropriate family based on data type
m_beta <- glmmTMB(rel.cc ~ temperature + time + (1 | PopulationPair), 
                  data = data, family = beta_family())

m_beta2 <- glmmTMB(rel.cc ~ temperature * time + (1 | PopulationPair), 
                  data = data, family = beta_family())

m_beta3 <- glmmTMB(rel.cc ~ temperature + (1 | PopulationPair), 
                  data = data, family = beta_family())

# Compare models based on AIC
aic_values <- data.frame(
  Model = c("m_beta", "m_beta2", "m_beta3"),
  AIC = c(AIC(m_beta), AIC(m_beta2), AIC(m_beta3))
)

# Print AIC values to determine the best model
print(aic_values)

# Identify the best model based on AIC
best_model <- aic_values[which.min(aic_values$AIC), ]
print(paste("The best model is:", best_model$Model, "with AIC:", best_model$AIC))

# Residual diagnostics for best model 
par(mfrow = c(1,2))  # Set up plot grid

# Plot residuals for Poisson model
residuals_best <- residuals(m_beta2, type = "pearson")
plot(residuals_best ~ fitted(m_beta2 ), 
     main = "Pearson Residuals vs Fitted (Beta)", xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

# Q-Q plot for residuals
qqnorm(residuals_best)
qqline(residuals_best, col = "red")


# Summarize the best model (for example, if m_neg_binom is the best)
summary(m_beta2)


# Tidy the model output for a cleaner view of fixed and random effects
tidy_model <- tidy(m_beta2)
print(tidy_model)


simulationOutput <- simulateResiduals(fittedModel = m_beta2, use.u = T)
plot(simulationOutput)
plot(simulationOutput, quantreg = TRUE)

testZeroInflation(simulationOutput)


# Check residuals
plot(residuals(m_beta2))
qqnorm(residuals(m_beta2))
qqline(residuals(m_beta2))
```

Generalized Linear Mixed Model (GLMM) - random effect Population -\> good fit less complicated

```{r}
###########################################
# 4. Generalized Linear Mixed Model (GLMM)#
###########################################
#----------------

# Fit different GLMMs with appropriate family based on data type
m_beta <- glmmTMB(rel.cc ~ temperature + time + (1 | Population), 
                  data = data, family = beta_family())

m_beta2 <- glmmTMB(rel.cc ~ temperature * time + (1 | Population), 
                  data = data, family = beta_family())

m_beta3 <- glmmTMB(rel.cc ~ temperature + (1 | Population), 
                  data = data, family = beta_family())

# Compare models based on AIC
aic_values <- data.frame(
  Model = c("m_beta", "m_beta2", "m_beta3"),
  AIC = c(AIC(m_beta), AIC(m_beta2), AIC(m_beta3))
)

# Print AIC values to determine the best model
print(aic_values)

# Identify the best model based on AIC
best_model <- aic_values[which.min(aic_values$AIC), ]
print(paste("The best model is:", best_model$Model, "with AIC:", best_model$AIC))

# Residual diagnostics for best model 
par(mfrow = c(1,2))  # Set up plot grid

# Plot residuals for Poisson model
residuals_best <- residuals(m_beta2, type = "pearson")
plot(residuals_best ~ fitted(m_beta2 ), 
     main = "Pearson Residuals vs Fitted (Beta)", xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

# Q-Q plot for residuals
qqnorm(residuals_best)
qqline(residuals_best, col = "red")


# Summarize the best model (for example, if m_neg_binom is the best)
summary(m_beta2)


# Tidy the model output for a cleaner view of fixed and random effects
tidy_model <- tidy(m_beta2)
print(tidy_model)


simulationOutput <- simulateResiduals(fittedModel = m_beta2, use.u = T)
plot(simulationOutput)
plot(simulationOutput, quantreg = TRUE)

testZeroInflation(simulationOutput)


# Check residuals
plot(residuals(m_beta2))
qqnorm(residuals(m_beta2))
qqline(residuals(m_beta2))
```

Time series accounted for in GLMM

```{r}
# Fit the Beta GLMM model
m_beta2 <- glmmTMB(rel.cc ~ temperature * time + (1 | PopulationPair), 
                   data = data, family = beta_family(),
                   control = glmmTMBControl(optimizer = optim,
                                            optArgs = list(method = "BFGS")),
                   dispformula = ~time)

# Residual diagnostics for best model 

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2)

# 1. Extract residuals
residuals_beta <- residuals(m_beta2)
fitted_values_beta <- fitted(m_beta2)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

par(mfrow = c(1,2))  # Set up plot grid
# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, residuals_beta,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(residuals_beta, main = "Q-Q Plot of Residuals")
qqline(residuals_beta, col = "red")

# 4. Plot Residuals vs Temperature (use aligned temperature)
plot(temperature_model, residuals_beta,
     xlab = "Temperature",
     ylab = "Pearson Residuals",
     main = "Residuals vs Temperature")
abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
plot(time_model, residuals_beta,
     xlab = "Time",
     ylab = "Residuals",
     main = "Residuals vs Time")
abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2)
sim_res2 <- simulateResiduals(fittedModel = m_beta2, group = data$PopulationPair)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)


summary(m_beta2)
```

Time series accounted for in GLMM - random effect Population --\> less complicated than best fit, this is the one I choose!!!

```{r}
# Fit the Beta GLMM model
m_beta2 <- glmmTMB(rel.cc ~ temperature * time + (1 | Population), 
                   data = data, family = beta_family(),
                   control = glmmTMBControl(optimizer = optim,
                                            optArgs = list(method = "BFGS")),
                   dispformula = ~time)

# Residual diagnostics for best model 

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2)

# 1. Extract residuals
residuals_beta <- residuals(m_beta2)
fitted_values_beta <- fitted(m_beta2)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

par(mfrow = c(1,2))  # Set up plot grid
# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, residuals_beta,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(residuals_beta, main = "Q-Q Plot of Residuals")
qqline(residuals_beta, col = "red")

# 4. Plot Residuals vs Temperature (use aligned temperature)
plot(temperature_model, residuals_beta,
     xlab = "Temperature",
     ylab = "Pearson Residuals",
     main = "Residuals vs Temperature")
abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
plot(time_model, residuals_beta,
     xlab = "Time",
     ylab = "Residuals",
     main = "Residuals vs Time")
abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2)
sim_res2 <- simulateResiduals(fittedModel = m_beta2, group = data$Population)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)


summary(m_beta2)


```

Quadratic effect on GLMM

-   <https://stats.stackexchange.com/questions/548478/how-to-interpret-meaning-of-residual-vs-predicted-quantile-plots-in-dharma?noredirect=1&lq=1>

-   <https://github.com/florianhartig/DHARMa/issues/285>

```{r}
# Remove rows with missing values in the relevant columns
clean_data <- na.omit(data[, c("rel.cc", "temperature", "time", "PopulationPair", "Population")])

# Fit the Beta GLMM model with a quadratic effect for temperature
# Including quadratic effects for temperature and time
m_beta2_quad <- glmmTMB(rel.cc ~ poly(temperature, 2) * poly(time, 2) + (1 | PopulationPair), 
                        data = clean_data, family = beta_family())


AIC(m_beta2)
AIC(m_beta2_quad) #  better than m_beta2!!!

summary(m_beta2_quad)

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2_quad)

# Check fitted values for NAs
fitted_values_beta <- fitted(m_beta2_quad)
sum(is.na(fitted_values_beta)) # Should return 0

sum(is.na(data$rel.cc))
sum(is.infinite(data$rel.cc))

# 1. Extract residuals
residuals_beta <- residuals(m_beta2)
fitted_values_beta <- fitted(m_beta2)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

par(mfrow = c(1,2))  # Set up plot grid
# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, residuals_beta,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(residuals_beta, main = "Q-Q Plot of Residuals")
qqline(residuals_beta, col = "red")

## 4. Plot Residuals vs Temperature (use aligned temperature)
#plot(temperature_model, residuals_beta,
#     xlab = "Temperature",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Temperature")
#abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
#plot(time_model, residuals_beta,
#     xlab = "Time",
#     ylab = "Residuals",
#     main = "Residuals vs Time")
#abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2)
sim_res2 <- simulateResiduals(fittedModel = m_beta2, group = data$PopulationPair)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)



```

Quadratic effect on GLMM - Population as random effect

-   <https://stats.stackexchange.com/questions/548478/how-to-interpret-meaning-of-residual-vs-predicted-quantile-plots-in-dharma?noredirect=1&lq=1>

-   <https://github.com/florianhartig/DHARMa/issues/285>

```{r}
# Remove rows with missing values in the relevant columns
clean_data <- na.omit(data[, c("rel.cc", "temperature", "time", "PopulationPair", "Population")])

# Fit the Beta GLMM model with a quadratic effect for temperature
# Including quadratic effects for temperature and time
m_beta2_quad <- glmmTMB(rel.cc ~ poly(temperature, 2) * poly(time, 2) + (1 | Population), 
                        data = clean_data, family = beta_family())


AIC(m_beta2)
AIC(m_beta2_quad) #  better than m_beta2!!!

summary(m_beta2_quad)

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2_quad)

# Check fitted values for NAs
fitted_values_beta <- fitted(m_beta2_quad)
sum(is.na(fitted_values_beta)) # Should return 0

sum(is.na(data$rel.cc))
sum(is.infinite(data$rel.cc))

# 1. Extract residuals
residuals_beta <- residuals(m_beta2)
fitted_values_beta <- fitted(m_beta2)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

par(mfrow = c(1,2))  # Set up plot grid
# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, residuals_beta,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(residuals_beta, main = "Q-Q Plot of Residuals")
qqline(residuals_beta, col = "red")

## 4. Plot Residuals vs Temperature (use aligned temperature)
#plot(temperature_model, residuals_beta,
#     xlab = "Temperature",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Temperature")
#abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
#plot(time_model, residuals_beta,
#     xlab = "Time",
#     ylab = "Residuals",
#     main = "Residuals vs Time")
#abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2)
sim_res2 <- simulateResiduals(fittedModel = m_beta2, group = data$Population)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)



```

Quadratic effect on GLMM and accounting for time series --\> best fit

```{r}
# Remove rows with missing values in the relevant columns
clean_data <- na.omit(data[, c("rel.cc", "temperature", "time", "PopulationPair", "Population")])

# Fit the Beta GLMM model with a quadratic effect for temperature
# Including quadratic effects for temperature and time
m_beta2_quad <- glmmTMB(rel.cc ~ poly(temperature, 2) * poly(time, 2) + (1 | PopulationPair), 
                 data = clean_data, family = beta_family(),
                 control = glmmTMBControl(optimizer = optim,
                                          optArgs = list(method = "BFGS")),
                 dispformula = ~time)

AIC(m_beta2)
AIC(m_beta2_quad) #  better than m_beta2!!!

summary(m_beta2_quad)

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2_quad)

# Check fitted values for NAs
fitted_values_beta <- fitted(m_beta2_quad)
sum(is.na(fitted_values_beta)) # Should return 0

sum(is.na(data$rel.cc))
sum(is.infinite(data$rel.cc))

# 1. Extract Pearson residuals
residuals_beta <- residuals(m_beta2_quad, type = "pearson")
manual_residuals <- clean_data$rel.cc - fitted(m_beta2_quad)
print(manual_residuals)

fitted_values_beta <- fitted(m_beta2_quad)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, manual_residuals,
     xlab = "Fitted Values",
     ylab = "Pearson Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(manual_residuals, main = "Q-Q Plot of Residuals")
qqline(manual_residuals, col = "red")

# 4. Plot Residuals vs Temperature (use aligned temperature)
#plot(temperature_model, manual_residuals,
#     xlab = "Temperature",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Temperature")
#abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
#plot(time_model, manual_residuals,
#     xlab = "Time",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Time")
#abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2_quad)
sim_res2 <- simulateResiduals(fittedModel = m_beta2_quad, group = data$Population)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)
```

Quadratic effect on GLMM and accounting for time series accounting for Population

```{r}
# Remove rows with missing values in the relevant columns
clean_data <- na.omit(data[, c("rel.cc", "temperature", "time", "PopulationPair", "Population")])

# Fit the Beta GLMM model with a quadratic effect for temperature
# Including quadratic effects for temperature and time
m_beta2_quad <- glmmTMB(rel.cc ~ poly(temperature, 2) * poly(time, 2) + (1 | Population), 
                 data = clean_data, family = beta_family(),
                 control = glmmTMBControl(optimizer = optim,
                                          optArgs = list(method = "BFGS")),
                 dispformula = ~time)

AIC(m_beta2)
AIC(m_beta2_quad) #  better than m_beta2!!!

summary(m_beta2_quad)

# Extract model data (rows used in the model)
model_data <- model.frame(m_beta2_quad)

# Check fitted values for NAs
fitted_values_beta <- fitted(m_beta2_quad)
sum(is.na(fitted_values_beta)) # Should return 0

sum(is.na(data$rel.cc))
sum(is.infinite(data$rel.cc))

# 1. Extract Pearson residuals
residuals_beta <- residuals(m_beta2_quad, type = "pearson")
manual_residuals <- clean_data$rel.cc - fitted(m_beta2_quad)
print(manual_residuals)

fitted_values_beta <- fitted(m_beta2_quad)

# Ensure temperature and time align with the residuals
temperature_model <- model_data$temperature
time_model <- model_data$time

par(mfrow = c(1,2))  # Set up plot grid
# 2. Plot Residuals vs Fitted Values
plot(fitted_values_beta, manual_residuals,
     xlab = "Fitted Values",
     ylab = "Pearson Residuals",
     main = "Residuals vs Fitted Plot")
abline(h = 0, col = "red")

# 3. Create a Q-Q plot for residuals
qqnorm(manual_residuals, main = "Q-Q Plot of Residuals")
qqline(manual_residuals, col = "red")

# 4. Plot Residuals vs Temperature (use aligned temperature)
#plot(temperature_model, manual_residuals,
#     xlab = "Temperature",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Temperature")
#abline(h = 0, col = "red")

# 5. Plot Residuals vs Time (use aligned time)
#plot(time_model, manual_residuals,
#     xlab = "Time",
#     ylab = "Pearson Residuals",
#     main = "Residuals vs Time")
#abline(h = 0, col = "red")

# 6. Use DHARMa for residual diagnostics
# Simulate residuals using DHARMa
sim_res <- simulateResiduals(fittedModel = m_beta2_quad)
sim_res2 <- simulateResiduals(fittedModel = m_beta2_quad, group = data$Population)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)
plot(sim_res, quantreg = FALSE)

plot(sim_res2)
plot(sim_res2, quantreg = FALSE)

# get residuals that are outside the simulation envelope
outliers(sim_res)
outliers(sim_res2)
```

ALternatives to GLMM

```{r eval=FALSE, include=FALSE}
# Using mgcv::gam for non-linear relationships

# Ensure PopulationPair is treated as a factor
data$PopulationPair <- as.factor(data$PopulationPair)

# Fit the GAM model with the beta family (logit link) and random effect
m_gam <- gam(rel.cc ~ s(temperature, k = 4) + s(time, k = 4) + s(PopulationPair, bs = "re"),
             family = betar(link = "logit"), data = data)

# Summary of the model
summary(m_gam)

# Diagnostic plots for the model
par(mfrow = c(2, 2))
gam.check(m_gam)

# Fit the previous GLMMTMB model to get the AIC value for comparison
m_beta2_quad <- glmmTMB(rel.cc ~ poly(temperature, 2) * poly(time, 2) + (1 | PopulationPair), 
                        data = clean_data, family = beta_family())
# Get AIC of the GLMMTMB model
aic_glmm <- AIC(m_beta2_quad)

# Fit a Generalized Additive Mixed Model (GAMM)
m_gamm <- gamm(rel.cc ~ s(temperature, k = 4) + s(time, k = 4), 
               random = list(PopulationPair = ~1), 
               family = gaussian(), 
               data = clean_data)

# Summarize the GAMM model
summary(m_gamm)

# Extract AIC for the GAM component
aic_gamm <- AIC(m_gamm$lme)

# Print the AIC
print(aic_gamm)

# Compare AIC values
cat("AIC of GLMMTMB model: ", aic_glmm, "\n")
cat("AIC of GAMM model: ", aic_gamm, "\n")

# Check for overfitting by visualizing the smooth terms
par(mfrow = c(1,2))  # Set up plotting area for two plots
plot(m_gamm$lme, pages = 1)  # Plot smooth terms

# Check residuals for GAMM
residuals_gamm <- residuals(m_gamm$lme)
fitted_values_gamm <- fitted(m_gamm$lme)

# Plot Residuals vs Fitted Values for GAMM
plot(fitted_values_gamm, residuals_gamm,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Plot for GAMM")
abline(h = 0, col = "red")

# Q-Q plot for residuals
qqnorm(residuals_gamm, main = "Q-Q Plot of Residuals for GAMM")
qqline(residuals_gamm, col = "red")

# Evaluate overfitting by checking if the residuals appear randomly scattered
# and if the fitted values adequately represent the data.


# Access the lme component
lme_summary <- summary(m_gamm$lme)
print(lme_summary)

# Access the gam component
gam_summary <- summary(m_gamm$gam)
print(gam_summary)

# Extract AIC from the GAM component
aic_gamm <- AIC(m_gamm$gam)
print(aic_gamm)


# Simulate residuals using DHARMa
#sim_res <- simulateResiduals(fittedModel = m_gamm$lme)
#sim_res2 <- simulateResiduals(fittedModel = m_beta2_quad, group = data$PopulationPair)

# 7. Plot DHARMa diagnostic plots
plot(sim_res)



library(brms)
#model <- brm(
#  rel.cc ~ s(temperature) + s(time) + (1 | PopulationPair),
#  family = Beta(),
#  data = clean_data,
#  autocor = cor_ar(~time | PopulationPair, p = 1) # Correct placement of autocorrelation
#)
duplicated_times <- any(duplicated(clean_data[, c("PopulationPair", "time")]))
if (duplicated_times) {
  print("There are duplicate time points within groups.")
}


# Check for colinearity
library(car)
vif(lm(rel.cc ~ temperature * time, data = data))
# there are higher-order terms (interactions) in this model


data$temperature_centered <- scale(data$temperature, center = TRUE, scale = FALSE)
data$time_centered <- scale(data$time, center = TRUE, scale = FALSE)

vif(lm(rel.cc ~ temperature_centered * time_centered, data = data))

# Example using mgcv package for regularization
m_gam <- gam(rel.cc ~ s(temperature, k = 4) + s(time, k = 4) + s(PopulationPair, bs = "re"),
             data = data, family = betar(link = "logit"))

summary(m_gam)

```
